---
title: "reverse_engineering_project"
output: html_document
date: "2024-04-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reverse Engineering Project

## Harper Lawson

### Background

I am examining the 2020 Washington Post story, “[Nearly 250 women have been fatally shot by police since 2015](https://www.washingtonpost.com/graphics/2020/investigations/police-shootings-women/)” by Marisa Iati, Jennifer Jenkins and Sommer Brugal.

An interactive database on the subject can be found [here](https://www.washingtonpost.com/graphics/investigations/police-shootings-database/).

This piece uses the murder of Breonna Taylor as a catalyst in diving into the much larger issue of police using excessive force disproportionally and unnecessarily against certain communities and demographics.

### Key Findings/Paragraphs

#### "Since 2015, Black women have accounted for less than 1 percent of the overall fatal shootings in cases where race was known. But within this small subset, Black women, who are 13 percent of the female population, account for 20 percent of the women shot and killed and 28 percent of the unarmed deaths."

I highlighted this paragraph because it is important to recognize how the percentages of Black women shot, killed, and killed while unarmed are higher than the percentage of the female population for which they account.

#### "Crenshaw said Black women’s deaths also may be dismissed as 'collateral damage' if they are killed while police are pursuing someone else. Twenty of the 247 women were killed in that kind of situation, analysis shows. In 12 of those 20 shootings, police said the women killed were caught in crossfire or shot accidentally."

I highlighted this paragraph because, in simple terms, it is a pretty big deal that 12 women were either killed in a crossfire or accidentally killed. While this might be a small percentage of the total deaths, it is an objectively high number for this matter of death.

#### "Since 2015, police have killed 26 unarmed women, including Taylor. Of those, 14 were White, seven were Black, four were Hispanic and one’s race was unknown. While about twice as many White women were shot and killed as Black women, White women account for five times the Black female population."

I highlighted this paragraph because of the key statistic that the population of White women is five times as much as that of Black women, but only twice as many White women were killed as Black women. This goes to show how the number of deaths of Black women in this case is still disproportionately high. Geoffrey P. Alpert, a criminology professor at the University of South Carolina and co-author of “Evaluating Police Uses of Force” was consulted about this information, likely due to the data analysis.

#### ["Of the 89 women killed at residences where they lived or often stayed, 12 encounters began as did Taylor’s — with a warrant to conduct a search or make an arrest in an investigation."

#### "In two killings, police said they shot the woman by accident. In most of those cases, the women were armed or reaching for a weapon, police said."]

I highlighted this paragraph because 12 killings beginning with a search or arrest warrant is, again, an objectively high number.

#### "About 31 percent, or 77, of the 247 women fatally shot by police since 2015 had mental health issues, compared to 22 percent of the 5,362 men killed."

I highlighted this paragraph because of the important statistic showing that more women with mental health issues (per the total) were killed than in the male population. John Snook, the Treatment Advocacy Center’s executive director was consulted about this information, likely due to the data analysis. The Treatment Advocacy Center is in Virginia and supports those struggling with mental health issues.

### First Steps

To load the necessary packages:

```{r}

library (tidyverse)
library (rmarkdown)
library (janitor)
library (ggplot2)
library (dplyr)
library (leaflet)
library (sf)

# Assuming these packages are already installed.
# Assuming base, datasets, forcats, graphics, purrr, readr, and stats are already installed and loaded.

```

To load the necessary data:

```{r}

shootings_data_2 <- read_csv("v2/fatal-police-shootings-data.csv")
shootings_agencies <- read_csv("v2/fatal-police-shootings-agencies.csv")

```

### Finding 1

#### "Since 2015, Black women have accounted for less than 1 percent of the overall fatal shootings in cases where race was known. But within this small subset, Black women, who are 13 percent of the female population, account for 20 percent of the women shot and killed and 28 percent of the unarmed deaths."

Filtering the data and obtaining values:

```{r}

# Filtering the data for fatal shootings where race is known
fatal_shootings_known_race <- subset(shootings_data_2, !is.na(race))

# Number of shootings where race is known
total_fatal_shootings_known_race <- nrow(fatal_shootings_known_race)

# Total number of Black women shot and killed
total_black_women_shot_and_killed <- nrow(subset(fatal_shootings_known_race, race == "B" & gender == "female"))

# Total number of all women shot and killed
total_women_shot_and_killed <- nrow(subset(fatal_shootings_known_race, gender == "female"))

# Total number of Black women shot and killed
total_black_women_shot_and_killed <- nrow(subset(fatal_shootings_known_race, race == "B" & gender == "female"))

# Total number of all unarmed people shot and killed
total_unarmed_shot_and_killed <- nrow(subset(fatal_shootings_known_race, armed_with == "unarmed"))

# Total number of all unarmed women shot and killed
total_unarmed_women_shot_and_killed <- nrow(subset(fatal_shootings_known_race, gender == "female" & armed_with == "unarmed"))

# Total number of unarmed Black women shot and killed
total_unarmed_black_women_shot_and_killed <- nrow(subset(fatal_shootings_known_race, race == "B" & gender == "female" & armed_with == "unarmed"))

```

"Black women have accounted for less than 1 percent of the overall fatal shootings in cases where race was known."

```{r}

# Percentage of Black women of all shot and killed
percentage_black_women_of_all <- total_black_women_shot_and_killed / total_fatal_shootings_known_race * 100

cat("Percentage of Black women out of overall fatal shootings in cases where race was known:", percentage_black_women_of_all, "%\n")

```

"Black women, who are 13 percent of the female population, account for 20 percent of the women shot and killed and 28 percent of the unarmed deaths."

```{r}

# Percentage of Black women out of all women shot and killed
percentage_black_women_of_all_women <- total_black_women_shot_and_killed / total_women_shot_and_killed * 100

# Percentage unarmed Black women out of all unarmed deaths
percentage_black_women_unarmed_deaths <- total_unarmed_black_women_shot_and_killed / total_unarmed_women_shot_and_killed * 100
    # *** WRONG

cat("Percentage of Black women of all women shot and killed:", percentage_black_women_of_all_women, "%\n")

cat("Percentage unarmed Black women shot and killed out of all unarmed deaths:", percentage_black_women_unarmed_deaths, "%\n")
    # *** WRONG

```

### Finding 2

#### "Crenshaw said Black women’s deaths also may be dismissed as 'collateral damage' if they are killed while police are pursuing someone else. Twenty of the 247 women were killed in that kind of situation, analysis shows. In 12 of those 20 shootings, police said the women killed were caught in crossfire or shot accidentally."

Obtaining values:

```{r}

# Total number of accidental deaths
total_accidental_deaths <- nrow(subset(fatal_shootings_known_race, threat_type == "accident" & gender == "female"))

total_accidental_deaths

```

### Finding 3

#### "Since 2015, police have killed 26 unarmed women, including Taylor. Of those, 14 were White, seven were Black, four were Hispanic and one’s race was unknown. While about twice as many White women were shot and killed as Black women, White women account for five times the Black female population."

Obtaining values:

```{r}

# Total number of all unarmed women shot and killed

total_unarmed_women_shot_and_killed

```

### Finding 4

#### ["Of the 89 women killed at residences where they lived or often stayed, 12 encounters began as did Taylor’s — with a warrant to conduct a search or make an arrest in an investigation."

#### "In two killings, police said they shot the woman by accident. In most of those cases, the women were armed or reaching for a weapon, police said."]

Obtaining values:

```{r}



```

### Finding 5

#### "About 31 percent, or 77, of the 247 women fatally shot by police since 2015 had mental health issues, compared to 22 percent of the 5,362 men killed."

Obtaining values:

```{r}



```
